Coupling finite and boundary element methods to solve the Poisson--Boltzmann equation for electrostatics in molecular solvation

### 2023-09-13 conclusion.tex
Updating according to Reviewers

This paper presents the first implementation of a FEM-BEM coupling approach to solve the Poisson--Boltzmann equation for molecular electrostatics. This brings the best of both worlds: the accuracy and efficiency of BEM to exactly enforce the boundary conditions at infinity, and the flexibility of FEM to account for space variations of the material properties and nonlinearities. After presenting verification results for a sphere and arginine with a constant permittivity inside the solute, we showcased our implementation with an advanced modeling technique that considers Gaussian-varying permittivities in a confined region, with the results validated against the widely-used APBS software. The final scaling results for larger molecules start from protein G B1 (955 atoms) and go up to immunoglobulin G (20\,148 atoms), proving the applicability of this approach for realistic problems. Even though our implementation was able to reach medium-to-large systems, we recongnize the need for further research towards better preconditioning of the linear system, optimizing the coupling technique, acceleration algorithms, and parallel execution, especially as we look towards much larger solutes, like viruses. We hope this proof-of-concept work will serve as motivation for future model development that considers space-varying permittivities and Debye lengths, and mixed linear-nonlinear techniques, especially for highly-charged systems, like nucleic acids.

>>>>

This paper presents the first implementation of a FEM-BEM coupling approach to solve the Poisson--Boltzmann equation for molecular electrostatics. This brings the best of both worlds: the accuracy and efficiency of BEM to exactly enforce the boundary conditions at infinity, and the flexibility of FEM to account for space variations of the material properties and nonlinearities. After presenting verification results of solvation energy for a sphere and arginine with a constant permittivity inside the solute, and binding energy for an extensive data set, we showcased our implementation with an advanced modelling technique that considers Gaussian-varying permittivities in a confined region, with the results validated against the widely-used APBS software. The final scaling results for larger molecules start from protein G B1 (955 atoms) and go up to immunoglobulin G (20\,148 atoms), proving the applicability of this approach for realistic problems. Even though our implementation was able to reach medium-to-large systems, we recognise the need for further research towards better preconditioning of the linear system, optimising the coupling technique, acceleration algorithms, and parallel execution, especially as we look towards much larger solutes, like viruses. We hope this proof-of-concept work will serve as motivation for future model development that considers space-varying permittivities and Debye lengths, and mixed linear-nonlinear techniques, especially for highly-charged systems, like nucleic acids.

### 2023-09-13 intro.tex
Updating according to Reviewers

In biologically relevant settings, the structure and function of biomolecules are largely determined by the surrounding water, which usually contains salt. 
To describe these systems accurately, we need to account for the solvent correctly, which has given rise to a wide range of models.\cite{onufriev2018water}
Highly detailed models consider every water molecule and salt ion explicitly.
For solutions with large numbers of molecules, however, these models can be very compuationally expensive, so implicit-solvent models---approximated models that use continuum theory to represent the ionic solution---are often used instead.\cite{RouxSimonson1999,DecherchiETal2015}
In the case of electrostatics, the implicit-solvent model is mathematically characterized by the Poisson--Boltzmann equation (PBE)\cite{Baker2004,Bardhan2012}, which is widely used to compute solvation free energies and mean-field potentials.

The implicit-solvent model for electrostatics describes the dissolved molecule as an infinite medium with a low-dielectric solute-shaped cavity, which contains a charge distribution from the partial charges---usually a sum of Dirac deltas at the atom's locations.
The outer solvent region is represented with a high dielectric constant, and considers the presence of salt.
These two regions are interfaced by the molecular surface where the continuity of the electrostatic potential and electric displacement are enforced.
The molecular surface can be defined in various ways \cite{HarrisBoschitcshFenley2013}.

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010}, boundary element\cite{boschitsch2002fast,LuETal2006,AltmanBardhanWhiteTidor09,bajaj2011efficient,GengKrasny2013,CooperBardhanBarba2014}, and analytical\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high-accuracy calculations \cite{GengKrasny2013,CooperBardhanBarba2014}, mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short: for example, if a variable permittivity is required inside the solute \cite{grant2001smooth,li2013dielectric}, or the solute is highly charged such that the linear approximation breaks\cite{FogolariETal1999}.

The present article describes a methodology to overcome some of those limitations, by coupling finite and boundary element methods.
This approach brings the best of both worlds---the flexibility of FEM and the efficiency of BEM---all in an accurate description of the solute molecule.
FEM-BEM coupling is a popular technique in the context of mixed linear-nonlinear models,\cite{carstensen1995coupling,aurada2013classical} fracture mechanics,\cite{aour2007coupled} fluid-structure interaction,\cite{estorff1991fem} acoustics,\cite{hiptmair2006stabilized} and electromagnetics.\cite{matsuoka1988calculation,hiptmair2008stabilized,bruckner20123d}
On the other hand, the PBE has been solved with hybrid numerical methods in the past: for example, by coupling finite differences with boundary elements\cite{boschitsch2004hybrid} or finite elements\cite{xie2016new,ying2018hybrid} to solve the nonlinear PBE in a specific region, and to implement modifications to the PBE model ({\it ie.} the size-modified PBE).
To the best of our knowledge, this is the first time finite and boundary elements have been combined in this application.

In this paper, we prototype this principle with the simplest implementation possible: a Johnson--N\'ed\'elec\cite{johnson1980coupling} coupling, where we solve using mass-matrix and diagonal preconditioning and whithout distributed memory execution.
This limits the size of problems we can access currently, however, it sets the basis for future developments that use more elaborate formulations and algorithms that are readily available in open source software libraries.
In this work, we use the boundary element library Bempp-cl\cite{BetckeScroggs2021} and the finite element library FEniCSx\cite{BasixJoss,BasixDofTransformations}.
These libraries are easy to use and their full functionalities may be accessed via their Python interfaces, making them ideal tools for easily implementing problems like this as well as for exploring computational efficiency and moving towards tackling large-scale problems, such as a full viral capsid.\cite{MartinezETal2019,wang2021high}
We hope this work will inspire research along these lines.

>>>>

In biologically relevant settings, the structure and function of biomolecules are largely determined by the surrounding water, which usually contains salt. 
To describe these systems accurately, we need to account for the solvent correctly, which has given rise to a wide range of models.\cite{onufriev2018water}
Highly detailed models consider every water molecule and salt ion explicitly.
For solutions with large numbers of molecules, however, these models can be very computationally expensive, so implicit-solvent models---approximated models that use continuum theory to represent the ionic solution---are often used instead.\cite{RouxSimonson1999,DecherchiETal2015}
In the case of electrostatics, the implicit-solvent model is mathematically characterized by the Poisson--Boltzmann equation (PBE)\cite{Baker2004,Bardhan2012}, which is widely used to compute solvation free energies and mean-field potentials.

The implicit-solvent model for electrostatics describes the dissolved molecule as an infinite medium with a low-dielectric solute-shaped cavity, which contains a charge distribution from the partial charges---usually a sum of Dirac deltas at the atom's locations.
The outer solvent region is represented with a high dielectric constant and considers the presence of salt.
These two regions are interfaced by the molecular surface where the continuity of the electrostatic potential and electric displacement are enforced.
The molecular surface can be defined in various ways \cite{HarrisBoschitcshFenley2013}.

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010,nakov2021argos}, boundary element\cite{boschitsch2002fast,LuETal2006,AltmanBardhanWhiteTidor09,bajaj2011efficient,GengKrasny2013,CooperBardhanBarba2014}, and analytic\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high-accuracy calculations \cite{GengKrasny2013,CooperBardhanBarba2014}, mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short: for example, if a variable permittivity is required inside the solute \cite{grant2001smooth,li2013dielectric}, or the solute is highly charged such that the linear approximation breaks\cite{FogolariETal1999}.

The present article describes a methodology to overcome some of those limitations, by coupling finite and boundary element methods.
This approach brings the best of both worlds---the flexibility of FEM and the efficiency of BEM---all in an accurate description of the solute molecule.
FEM-BEM coupling is a popular technique in the context of mixed linear-nonlinear models,\cite{carstensen1995coupling,aurada2013classical} fracture mechanics,\cite{aour2007coupled} fluid-structure interaction,\cite{estorff1991fem} acoustics,\cite{hiptmair2006stabilized} and electromagnetics.\cite{matsuoka1988calculation,hiptmair2008stabilized,bruckner20123d}
On the other hand, the PBE has been solved with hybrid numerical methods in the past: for example, by coupling finite differences with boundary elements\cite{boschitsch2004hybrid} or finite elements\cite{xie2016new,ying2018hybrid} to solve the nonlinear PBE in a specific region, and to implement modifications to the PBE model (i.e. the size-modified PBE).
To the best of our knowledge, this is the first time finite and boundary elements have been combined in this application.

In this paper, we prototype this principle with the simplest implementation possible: a Johnson--N\'ed\'elec\cite{johnson1980coupling} coupling, where we solve using mass-matrix and diagonal preconditioning and without distributed memory execution.
This limits the size of problems we can access currently, however, it sets the basis for future developments that use more elaborate formulations and algorithms that are readily available in open-source software libraries.
In this work, we use the boundary element library Bempp-cl\cite{BetckeScroggs2021} and the finite element library FEniCSx\cite{BasixJoss,BasixDofTransformations}.
These libraries are easy to use and their full functionalities may be accessed via their Python interfaces, making them ideal tools for easily implementing problems like this as well as for exploring computational efficiency and moving towards tackling large-scale problems, such as a full viral capsid.\cite{MartinezETal2019,wang2021high}
We hope this work will inspire research along these lines.

### 2023-09-13 main.tex
Updating according to Reviewers

\begin{abstract}
The Poisson--Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piecewise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson--Boltzmann equation, where the finite element method is applied in a confined {\it solute} region, and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we use the simplest methods available: Johnson--N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. We showcase our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. As validation, we compare against well-established finite difference solvers for an extensive binding energy data set, and with the finite difference code APBS (to 0.5\%) for Gaussian permittivities. We also show scaling results from protein G B1 (955 atoms) up to immunoglobulin G (20\,148 atoms). For small problems, the coupled method was efficient, outperforming a purely boundary integral approach. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements alone, we were able to run medium to large sized problems on a single workstation. Development of better preconditioning techniques and the use of distributed memory parallelism for larger systems remains an area for future work. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

@@

%((Main text paragraphs should be 12 point font, double-spaced. Reviews are comprehensive survey of recent progress in a topic of broad interest in quantum chemistry, providing the readership with an appreciation of the importance of the work, a summary of recent developments, and a guide to the relevant literature. Perspective are short discussions of an important emerging topics in quantum chemistry, usually focused on no more than a few recently published papers, and including the authors' vision for the future of the topics, identifying important problems that should be addressed next. Perspective should be limited to 3000 words, 4 display items (figures and/or tables), and 30 references.))

@@

\subsection*{\sffamily \large ACKNOWLEDGMENTS}
MB acknowledges the support from Kingston University through First Kingston University Grant. \\
MS acknowledges support from EPSRC grant EP/W007460/1 \\
TB acknowledges support from EPSRC grants EP/W007460/1 and EP/W026260/1. \\
EB acknowledges support from EPSRC grants EP/V050400/1 and EP/W007460/1. \\
CDC acknowledges the support from CCTVal through ANID PIA/APOYO AFB220004.\\
We would like to thank J{\o}rgen Dokken for his help with various implementation details related to FEniCSx.
%((Place Acknowledgments here))

>>>>

\begin{abstract}
The Poisson--Boltzmann equation is widely used to model electrostatics in molecular systems. Available software packages solve it using finite difference, finite element, and boundary element methods, where the latter is attractive due to the accurate representation of the molecular surface and partial charges, and exact enforcement of the boundary conditions at infinity. However, the boundary element method is limited to linear equations and piecewise constant variations of the material properties. In this work, we present a scheme that couples finite and boundary elements for the Poisson--Boltzmann equation, where the finite element method is applied in a confined {\it solute} region and the boundary element method in the external {\it solvent} region. As a proof-of-concept exercise, we use the simplest methods available: Johnson--N\'ed\'elec coupling with mass matrix and diagonal preconditioning, implemented using the Bempp-cl and FEniCSx libraries via their Python interfaces. We showcase our implementation by computing the polar component of the solvation free energy of a set of molecules using a constant and a Gaussian-varying permittivity. As validation, we compare against well-established finite difference solvers for an extensive binding energy data set, and with the finite difference code APBS (to 0.5\%) for Gaussian permittivities. We also show scaling results from protein G B1 (955 atoms) up to immunoglobulin G (20\,148 atoms). For small problems, the coupled method was efficient, outperforming a purely boundary integral approach. For Gaussian-varying permittivities, which are beyond the applicability of boundary elements alone, we were able to run medium to large-sized problems on a single workstation. The development of better preconditioning techniques and the use of distributed memory parallelism for larger systems remains an area for future work. We hope this work will serve as inspiration for future developments that consider space-varying field parameters, and mixed linear-nonlinear schemes for molecular electrostatics with implicit solvent models. 
\end{abstract}

@@

%((Main text paragraphs should be 12 point font, double-spaced. Reviews are a comprehensive survey of recent progress in a topic of broad interest in quantum chemistry, providing the readership with an appreciation of the importance of the work, a summary of recent developments, and a guide to the relevant literature. Perspective are short discussions of an important emerging topics in quantum chemistry, usually focused on no more than a few recently published papers, and including the authors' vision for the future of the topics, identifying important problems that should be addressed next. Perspective should be limited to 3000 words, 4 display items (figures and/or tables), and 30 references.))

@@

\subsection*{\sffamily \large ACKNOWLEDGMENTS}
MB acknowledges the support from Kingston University through the First Kingston University Grant. \\
MS acknowledges support from EPSRC grant EP/W007460/1 \\
TB acknowledges support from EPSRC grants EP/W007460/1 and EP/W026260/1. \\
EB acknowledges support from EPSRC grants EP/V050400/1 and EP/W007460/1. \\
CDC acknowledges the support from CCTVal through ANID PIA/APOYO AFB220004.\\
We would like to thank J{\o}rgen Dokken for his help with various implementation details related to FEniCSx.
%((Place Acknowledgments here))
